---#David Landerman, Jour772, pre_lab_01 - data_journalism_book_chapter_submission

title: "pre_lab_01.Rmd"
author: "Derek Willis / Rob Wells"
date: "2025-09-09"
output: html_document
---
#Update to fix confusing question

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## About this notebook

This notebook contains code and explanatory text that your should review and run as you read through chapters 13 and 15 of the course textbook, "Data Journalism with R and the Tidyverse". https://thescoop.org/data_journalism_book/
Answer questions and edit the document as directed.

Running this notebook will help you understand key data analysis methods and concepts that you will put into practice during this week's lab. When you come across a new command or verb in R, you should add it to your class reference notebook file along with what it's used for.

When you are finished running the code in this notebook, you will push changes to your course GitHub repo, and upload the link to ELMS as instructed.

You also should update your class reference notebook with code examples and what they mean. When you come across a new concept or R verb, put it in your reference notebook along with a sentence about what it does. For example:

```{r}
library(tidyverse)
```

The library function loads an R library such as the tidyverse, providing access to the verbs and functions that library contains. You have to load a library before you can use its functions.

## Instructor in class demonstration outline

-   Load this notebook, and provide overview of R, RStudio (text editor, console, environment window, file explorer), Rmarkdown data notebooks, the reference notebook.
-   Work through the code below, chapter by chapter.
    -   Chapter 13: explain the console and the concept of variables. Explain packages, the Tidyverse, how to install them.
    -   Chapter 15: explain loading libraries, the concept of functions, loading data, the importance of having questions to guide data analysis, and aggregation and summarization.

## Chapter 13

### Task 1: The console

**Task** In the console (look below), type 2+2 and hit enter. What does it return? Type your answer below. 
**Answer:[1] 4**

### Task 2: Variables

**Task** Put the following code -- which stores the numeral 2 as a variable called number -- in the console and hit enter.

```{r}
number <- 2
```

Then type this code to multiply the variable called number by itself.

```{r}
number * number
```

What does it return? Type your answer below:

**Answer:** [1] 4

### Task 3: Install packages

**Task** In the console, run the following code to install the Tidyverse:

```{r}
install.packages('tidyverse')
```

**Task** Add that

Then run the following code to install the rmarkdown library

```{r}
install.packages('rmarkdown')
```

### Task 4: Load libraries

**Task** Run the following code in the gray-colored codeblock below -- not in the console -- to load the tidyverse library. To run the code, click the little green play button (left facing arrow) at the top right of the codeblock. In Rmarkdown data notebooks, we write code inside of codeblocks, and explanatory text in the white area outside of it.

```{r}
# This is a comment inside of a codeblock. Comments doesn't run when we run the codeblock.
# Adding this command is a good way to start your class reference notebook.
library(tidyverse)
library(rmarkdown)
```

## Chapter 15

### Task 5: Load data

**Task** Load the UMD courses data by running the following codeblock. The data is contained in an .rds file, which is a binary format for R data.

```{r}
umd_courses <- read_rds("umd_courses.rds")
```

Once it loads, look at the environment window at right in RStudio for umd_courses It's next to a small blue circle with a white arrow in it. To the right of that, look at how many "obs" the dataset has ("obs" is short for observation, which is another way of saying rows or records). Look at how many variables it has (another way of saying columns or fields). Write the number of observations and variables below.

**Answer**
79366
#dl comment don't ding me for not generating ^this number with code, IT SAYS "LOOK" and "write"

### Task 6: Glimpse data

**Task** Run the following codeblock to use glimpse() to get a sense of the data, some of the values in each field/variable/column, and the data types of each field/variable/column.

```{r}
glimpse(umd_courses)
```

In the space below, write the data type for the "id" column and the "seats" column (it's in light gray, offset by \< \>). **Answer** id column: chr, seats column: dbl

### Task 7: Examine data "head"

**Task** Run the following codeblock to use head() to examine the first six rows of data. Explore the data by clicking the black arrow on the right side of the output table, allowing you to examine additional columns.

```{r}
head(umd_courses)
```

#Update:
In the space below, write the name of the entry from the first row of the "title" column. 
 **Answer** Directed Research

### Task 8: Explore data with interactive dataframe

**Task** look at the environment window at right in RStudio for umd_courses. It's next to a small blue circle with a white arrow in it. Click on the umd_courses -- not the blue circle. It will open the dataframe as an interactive table. Scroll down to see additional rows. Scroll left to see additional columns. Click on a column name to sort by that column.

### Task 9: Group and count

So what if we wanted to know how many classes were offered in each term?

To do that by hand, we'd have to take each of the 79,366 individual rows (or observations or records) and sort them into a pile. We'd put them in groups -- one for each term -- and then count them. To do this, we'll take our dataset and we'll introduce a new operator: `|>`. The best way to read that operator, in my opinion, is to interpret that as "and then do this." The keyboard shortcut for this is cmd-shift-m on the Mac and ctrl-shift-m on Windows. Don't like that character? R now has a new one that does the same thing: `|>`.

**Task** Run the following codeblock:

```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n()
  )
```

Describe what you think each line of this code is doing below, based on the result:

**Answer**
Ok the question says below so BASED ON WHAT WAS ABOVE, the count classes in each term, I am going to guess what is going on BELOW is: doing the exact same thing as above except its going to arrange the rows by terms with the most classes first and then descending order.

### Task 10: Adding arrange to sort the data

**Task** Run the following codeblock:

```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(desc(count_classes))
```

Compared to the previous codeblock, what has changed and how did that affect the result? 
**Answer** arrange(desc(count_classes)) was added to the block, which arranges the count of classes in descending order

### Task 11: Grouping by multiple columns

**Task** Run the following codeblock.

```{r}
umd_courses |>
  group_by(term, department) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(term, department)
```

Compared to the previous codeblock, how did the output change and what in the code changed it?

**Answer** Ok this is weird there are two little weird windows in the R markdown one says "R console" and the other says "grouped_df" under that it says "A tribble 4270x3" (isn't tribble a star trek thing) and "Groups term [29]" what is that nonsense about?
Anyways my answer is that it ordered the original table by first term then within those terms title, I AM GUESSING by looking at the lists that it defaults to alphabetical

### Task 12: Other summary functions

**Task** Run the following codeblock:

```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats)
  ) |>
  arrange(desc(total_seats))
```

Compared to the previous codeblock, what has changed and how did that affect the result? **Answer**So its totalling the number of classes in each term and also how many total seats in each term and arranging the table by descending value of total seats

### Task 13: different kinds of summarizations

**Task** Run the following codeblock to group the umd_courses dataframe by term. Then do the following summarizations for each term: count the number of courses, add up the total of all seats, and calculate the mean and median contribution number of seats. Then sort so the term with the most total seats is listed first.

```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    mean_seats = mean(seats),
    median_seats = median(seats)
  ) |>
  arrange(desc(total_seats))
```

Compared to the previous codeblock, what has changed and how did that affect the result? What do the top terms have in common? **Answer**So this basically added two columns and I was confused about what data those mean and median operations were working on so apparently its the mean/median number of seats in each...class? Ya because thats the only data set within each term...

### Task 14: more aggregations

**Task** Run the following codeblock to group the umd_courses dataframe by the department. Then do the following summarizations for each group: count the number of courses, add up the total of all seats, calculate the mean and median number of seats and then add lines to calculate the smallest and largest seat numbers using the `min()` and `max()` functions. Then sort so the department with the largest total number of seats is listed first.

```{r}
umd_courses |>
  group_by(department) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    mean_seats = mean(seats),
    median_seats = median(seats),
    min_seats = min(seats),
    max_seats = max(seats)
  ) |>
  arrange(desc(total_seats))
```

Compared to the previous codeblock, what has changed and how did that affect the result? How is the data sorted?

**Answer** Well in this you're grouping by department and not term, and you're finding basically how many classes and seats available in those classes. Median and mean are...basic functions

### Task 12: Questions

No code here; I want you to provide three questions that this data suggests to you. What would you like to ask it?

**Answer** This is so open ended I dont know if you mean specifically operation wise or if I am asking questions story wise, qualitatively and it after 9pm so I cant email you.

1.  Well for starters I would want to know what engineering science is, very strange department title. How about how many classes each professor teaches

2.  How many sections in total a professor teaches

3.  How many students those instructors have the opportunity to scare the bejesus out of
